Inference of a universal social scale and segregation measures using social connectivity kernels

### 2020-09-30 manuscript/main.tex
Submitted version.

\documentclass{scrartcl}
\usepackage{authblk}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{csquotes}
\usepackage[british]{babel}
\usepackage[
    giveninits,
    backend=biber,
    sorting=none,
    citestyle=numeric-comp,
    maxnames=3,
]{biblatex}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{cleveref}
\usepackage{fmtcount}
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf}

@@

@@

    We use the connectivity kernel of conditionally-independent edge models to develop a family of segregation statistics with desirable properties: they offer an intuitive and universal characteristic scale on social space (facilitating comparison across datasets and societies), are applicable to multivariate and mixed node attributes, and capture segregation at the level of individuals, pairs of individuals, and society as a whole. We show that the segregation statistics induce a (semi-)metric on Blau space (a space spanned by the attributes of the members of society) and provide maps of two societies.

@@

\section{Introduction\label{sec:introduction}}

@@

Homophily can be observed in friendships~\cite{Currarini2009, Hipp2009}, networks of discussion partners~\cite{McPherson2006}, communication networks~\cite{Wang2013, Leo2016}, marital ties~\cite{Blau1984}, and online social networks~\cite{Chang2010}. Relationships are homogeneous with respect to a wide range of attributes, including age~\cite{Marsden1988,Smith2014}, sex~\cite{Stehle2011,Smith2014}, ethnicity~\cite{Chang2010, Blumenstock2013, Currarini2009}, education~\cite{McPherson2006, Smith2014,Johnson1989}, occupation~\cite{Chan2004}, income~\cite{Leo2016, Wang2013, Johnson1989}, religion~\cite{Platt2012}, parental~\cite{Johnson1989} and marital status~\cite{Kalmijn2007}, political ideology~\cite{Bakshy2015, Boutyline2017}, and geographical location~\cite{Lambiotte2008, Expert2011, Backstrom2010, Scellato2011, Illenberger2013}.

Segregation statistics, also referred to as segregation measures, are often used to quantify homophily~\cite{Rodriguez-Moral2016, Bojanowski2014}. Many approaches are based on co-presence in organisational units such as schools~\cite{Orfield2014}, voluntary associations~\cite{Popielarz1999}, occupations~\cite{Charles1995}, or census tracts~\cite{Reardon2004,Reardon2011}, and we refer to them as \emph{organisational} statistics. Typically, they compare how the distribution of demographic attributes within organisational units differs from the distribution of attributes in the general population. Whilst organisational statistics are applicable whenever data can be stratified according to a variable of interest, they cannot capture segregation at smaller scales than the strata~\cite{Blumenstock2013}. For example, the ethnic composition of a set of schools may be representative of the general population, indicating that there is no (organisational) segregation. But the social networks \emph{within} schools often exhibit strong ethnic homophily~\cite{Currarini2009,Moody2001}. Organisational statistics cannot capture such social segregation.

@@

Building on the successes of conditionally-independent edge models~\cite{Snijders2011} and, in particular, latent space models for social networks~\cite{Hoff2002,Handcock2007,Raftery2012}, we consider a generative model for social networks whose members occupy a multidimensional Blau space in \cref{sec:model}. We discuss desirable properties for social segregation statistics, and, using the generative network model, we develop a suite of statistics applicable to arbitrary attributes in \cref{sec:model-based-segregation}. The statistics capture segregation at different scales: single individuals, pairs of individuals, and society as a whole. Because of both their probabilistic foundations and their construction from the universal notion of a social tie, the segregation statistics have a universal scale, i.e.\ one unit of segregation has the same implications across different societies or at different times. We show that the segregation statistic for pairs of individuals is a semi-metric and can be used to quantify distance in Blau space. We illustrate the statistics with a simple example, and we show that it reduces to a well-known segregation statistic if the attributes are univariate and categorical: the natural logarithm of \citeauthor{Moody2001}'s $\alpha$ index~\cite{Moody2001}.

@@

We consider a generative model for social networks for a population of $n$ individuals $\population$ who occupy a Blau space $\blauspace$ spanned by their demographic attributes, such as age, income, or sex. In contrast to common latent space models~\cite{Hoff2002, Hoff2008, Handcock2007}, the attributes are observed, although \textcite{Hoff2002} also consider an extension including covariates. The $q$-dimensional attribute vector $x_i\in\blauspace$ for each individual $i\in \population$ is drawn independently from a distribution $P(x_i)$ of demographic attributes. Elements of the attribute vector can take continuous, ordinal, or categorical values. Connections between individuals are encoded by the binary adjacency matrix $A$ such that $A_{ij}=1$ if $j$ considers $i$ to be a friend and $A_{ij}=0$ otherwise. We assume that people do not interact with themselves such that $A_{ii}=0$ for all $i$, and that connections are undirected, although social ties need not be reciprocated in general~\cite{Ball2013}.

@@

We have so far emphasised the desirability of metrics on social spaces with a universal characteristic scale (in the sense of being comparable between societies). After first developing social segregation statistics, including a notion of social separation, we will formulate both a metric and a notion of scale in \cref{sec:segregation-distance}.

@@

In other words, the social separation statistic is a true measure of \emph{distance} in the social space with a probabilistic interpretation if features are themselves measures of distance, including all the features we consider subsequently. This observation puts Peter Blau's~\cite{Blau1977} hypothesis that ``the macrostructure of societies can be defined as a multidimensional space of social positions among which people are distributed and which affect their social relations'' on a sound statistical footing: \emph{fitting conditionally-independent edge models allows us to learn the metric of Blau space}. The metric has a universal scale: \emph{one unit of social separation has the same probabilistic meaning independent of the society under consideration}, facilitating comparison across disparate datasets. Even if two societies have different Blau space dimensions, e.g.\ a society might exist that strongly discriminates based on characteristics which are not found in other societies, the social separation between a pair of individuals has a common meaning. However, it is worth noting that we cannot quantify segregation due to unobserved dimensions of the social space.

@@

The General Social Survey (GSS) is a nationally-representative face-to-face survey of non-institutionalised adults living in the US. Demographic attributes of seeds are collected regularly and include age, sex, ethnicity, religion, and education~\cite{Marsden1987,Marsden1988}. In 2004, respondents were asked about the demographic background of people ``with whom they discuss important matters'', which tends to elicit close ties~\cite{Marin2004}. We omit all nominees who are not considered to be friends or who are family. Some of the demographic attributes of seeds and nominees are missing because respondents did not know or refused to provide the information, and we drop dyads associated with individuals with one or more missing attributes, as shown in \cref{tbl:survey-sample-size}. Such a complete-case analysis can introduce biases if the data are not missing completely at random~\cite{Rubin1976}, but handling the missing data in a principled fashion would require us to develop a model for demographic attributes~\cite{Pigott2001, Gelman2013}.

The coding of age and sex is consistent amongst seeds and nominees. We aggregate the detailed coding of ethnic and religious attributes of seeds to match the coding of nominees, as shown in \cref{tbl:survey-gss-coding}. Kernel features include the absolute age and ordinal education level difference as well as binary indicators for differences along the sex, ethnicity, and religion dimensions. For each demographic attribute, we define a feature for the logistic kernel in \cref{eq:logistic-kernel}, as shown in \cref{tbl:survey-gss-coding}. To standardise the features $f(x_i,x_j)$, we subtract their mean and divide non-binary features by twice their standard deviation~\cite{Gelman2008a}; binary features are not rescaled. The statistics are calculated with respect to a random sample of pairs of seeds. Feature standardisation allows us to compare kernel parameters more easily~\cite{Gelman2008a} and simplifies the formulation of priors: we use independent, weakly-informative Cauchy priors for the kernel parameters such that
\[
    P(\theta_l) \propto {\left[1+{\left(\frac{\theta_l}{\alpha_l}\right)}^2\right]}^{-1}.
\]
Following \textcite{Gelman2008}, we chose the scale parameters $\alpha_l=2.5$ for $l>1$ to represent our weak prior belief that changing a feature by one standard deviation is unlikely to change the log odds by more than five: the independent Cauchy distributions regularise the kernel parameters by placing significant prior probability near zero, but their heavy tails allow for significant departures from zero should the data be in support of large parameters. We set $\alpha_1=10$ because the parameter $\theta_1$ associated with the constant bias term could change significantly depending on the population size~\cite[chapter 16]{Gelman2013}.

@@

To get a better understanding of Blau space and the metric induced by the connectivity kernel, we consider a sample $S$ of 1,000 respondents from the GSS and USS. For each sample, we compute the social separation between pairs of respondents to obtain a distance matrix
\[
    \hat\varphi_{ij} = \transpose{\hat\theta}\left(f(x_j, x_j)-f(x_i, x_j)\right),
\]
where $\hat\theta$ is the posterior median of the kernel parameters discussed in \cref{sec:survey-gss,sec:survey-bhps-usoc}. We use multidimensional scaling to embed the respondents in a two-dimensional space~\cite{Borg1996}, as shown in \cref{fig:segregation-embedding}. Panels~(c) and~(d) show the two-dimensional embedding that best approximates the distance matrix in the high-dimensional social space (we omit contribution due to physical space for the USS to make the embeddings comparable).

@@

Panel~(c) and~(d) of \cref{fig:segregation-embedding} also show the social isolation $\phi$ experienced by individuals as a greyscale heat map which we obtained in two steps: first, we evaluated an estimate of the social isolation
\[
    \hat\phi_i = \frac{1}{\card{S}-1}\sum_{j\in S:j\neq i} \hat\varphi_{ij}.
\]
Second, we applied Gaussian kernel smoothing to the social isolation in the embedding space. Respondents occupying the centre of society experience little isolation whereas individuals in the periphery are more isolated. For example, members of the ethnic majority experience an average social isolation of 4.53 (4.47--4.59 95\% credible interval) in the USS and 4.46 (4.10--4.84 95\% credible interval) in the GSS. In contrast, the average social isolation amongst ethnic minorities is 4.99 (4.91--5.07 95\% credible interval) and 5.16 (4.73--5.58 95\% credible interval): significantly higher than for the ethnic majority.

@@

Similar to the social separation in \cref{eq:social-separation-logistic}, the social strain can be broken down into components
\begin{equation}
    \Phi_l = \theta_l \int dx\ dy\ \left(f_l(y, y) - f_l(x, y)\right)\label{eq:segregation-contributions}
\end{equation}
because it is a linear functional of $\varphi$: each component contributes to the social strain in society. For each of the datasets, we evaluate an estimate of the contributions to the social strain
\[
    \hat\Phi_l = \frac{2\theta_l}{\card{\seeds}(\card{\seeds}-1)} \sum_{i < j\in\seeds:i\neq j} \left[f_l(x_j, x_j)-f_l(x_i, x_j)\right],
\]
where the sum is over all distinct pairs of seeds $\seeds$. The contribution $\Phi_l$ quantifies the average social separation due to feature $l$, and it captures the effect of both the connectivity kernel $\rho(x,y)$ and the attribute distribution $P(x)$ in Blau space: neither is sufficient on its own to quantify segregation. Furthermore, $\Phi$ and its contributions in \cref{eq:segregation-contributions} can facilitate comparison across different datasets, as illustrated in \cref{fig:segregation-embedding}: they have an intuitive interpretation (the probability of edges decreases with increasing segregation) and universal scale (one unit of segregation has the same effect on edge probability).

@@

Even though we did not expect the kernel parameters to be consistent across countries, time, or even different surveys, people connected with one another in a surprisingly similar fashion across the different datasets (the BHPS and USS are longitudinal studies such that consistent parameter estimates are less surprising). Our observations, together with a study by \textcite{Mossong2008} finding that ``mixing patterns [\ldots] were remarkably similar across different European countries'', suggest that connectivity kernels for friendships vary little across societies and time. To test this hypothesis, further surveys should be conducted in a unified fashion to minimise the effects of question wording and how the survey is administered~\cite{Eagle2015}. In particular, such surveys should explore options to explicitly incentivise nominees to provide data about themselves~\cite{Biernacki1981}: seeds may not recall certain attributes, or nominees may deliberately portray themselves inaccurately~\cite{Bruch2016}. Questions regarding ethnicity should allow respondents to provide multiple answers such that people with mixed ethnic backgrounds can express their identity. Rather than asking respondents about potentially sensitive information, such as income, proxy information that is more readily available---and potentially more informative of how individuals interact with society---could be collected~\cite{Po2012}. Whenever possible, aggregation of attributes such as age into bins should be avoided because it limits the ability to infer kernel parameters~\cite{Hutcheon2010}, as we saw in \cref{sec:survey-alp}.

@@

We would like to thank Sahil Loomba for useful feedback on the manuscript.

\printbibliography

@@

\section{Evaluation of the observed-data log-likelihood}

@@

In the first step, we draw the positions $x$ of $n=2,000$ nodes uniformly at random from the unit square, and we connect nodes to one another according to a logistic connectivity kernel with features
\[
    f(x_i, x_j) = \left(1, \frac{3 \left[\abs{x_{i1}-x_{j1}} - \frac{1}{3}\right]}{\sqrt{2}}, \frac{3 \left[\abs{x_{i2}-x_{j2}} - \frac{1}{3}\right]}{\sqrt{2}}\right),
\]
where the first feature represents the bias, and the last two features capture distance in Blau space. The features were chosen to be standardised in the same fashion as described in \cref{sec:survey-gss}, i.e.\ to have zero mean and a standard deviation of $0.5$~\cite{Gelman2008a}. The corresponding parameter values $\theta$ are drawn from a normal distribution with unit variance and expectation
\[
    \left\langle \theta\right\rangle = \left(-7, 0, 0\right).
\]
The expectation $\langle\theta\rangle$ was chosen such that it gives rise to a typical degree of $2,000 \times \expit(-7)\approx 1.8$, similar to the real-world datasets considered in \cref{sec:application}. We select $s=100$ respondents as egos and include all their alters as positive examples. We sample three times as many negative examples by selecting distinct pairs of respondents. If the number of respondents is not sufficient to draw the desired number of distinct control pairs, we use all $\frac{s(s-1)}{2}$ possible distinct pairs of respondents as negative examples.

@@

In the USS, respondents could identify with mixed ethnicities, and we coded such responses as a mixed membership. For example, a respondent who indicated ``mixed Asian and White'' would belong to both White and Asian ethnicities. To quantify how different two people are in terms of ethnicity, we define the feature map
\[
    f_\text{ethnicity}(x_i, x_j) = \frac{1}{2} \sum_{l\in E}\abs{x_{il}-x_{jl}},
\]
where $E$ is the set of attributes encoding ethnic identity, and ethnicity memberships are normalised such that $\sum_{l\in E}x_{jl}=1$ for all $j$. For example, $f_\text{ethnicity}(x_i, x_j)=1$ for two people $i$ and $j$ one of which identifies as white and the other as black. For a person $i$ identifying as black and another person $j$ identifying as mixed black and Asian, $f_\text{ethnicity}(x_i, x_j)=0.5$.

@@

>>>>

\documentclass{scrartcl}
\usepackage{authblk}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{csquotes}
\usepackage[british]{babel}
\usepackage[
    giveninits,
    backend=biber,
    sorting=none,
    citestyle=numeric-comp,
    maxnames=10,
    style=vancouver,
]{biblatex}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{cleveref}
\usepackage{fmtcount}
\usepackage{environ}
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf}

@@

\def\composite{}
% Useful for word counting
\ifx\composite\undefined
  % hide graphics
  \renewcommand{\includegraphics}[1]{}
  % hide pagenumber
  \pagestyle{empty}
  \renewcommand*{\titlepagestyle}{empty}
  % hide equations
  \RenewEnviron{equation}{equation environment}
  \RenewEnviron{align}{align environment}
  \RenewEnviron{equation*}{equation* environment}
\fi

@@

    We use the connectivity kernel of conditionally-independent edge models to develop a family of segregation statistics with desirable properties: they offer an intuitive and universal characteristic scale on social space (facilitating comparison across datasets and societies), are applicable to multivariate and mixed node attributes, and capture segregation at the level of individuals, pairs of individuals, and society as a whole. We show that the segregation statistics can induce a metric on Blau space (a space spanned by the attributes of the members of society) and provide maps of two societies.

@@

\begin{refsection}
\section{Introduction\label{sec:introduction}}

@@

Homophily can be observed in friendships~\cite{Currarini2009, Hipp2009}, networks of discussion partners~\cite{McPherson2006}, communication networks~\cite{Wang2013, Leo2016}, marital ties~\cite{Blau1984}, and online social networks~\cite{Chang2010}. Relationships are homogeneous with respect to a wide range of attributes, including age~\cite{Marsden1988,Smith2014}, sex~\cite{Smith2014}, ethnicity~\cite{Chang2010, Blumenstock2013, Currarini2009}, education~\cite{McPherson2006, Smith2014,Johnson1989}, occupation~\cite{Chan2004}, income~\cite{Leo2016, Wang2013, Johnson1989}, religion~\cite{Platt2012}, parental~\cite{Johnson1989} and marital status~\cite{Kalmijn2007}, political ideology~\cite{Bakshy2015, Boutyline2017}, and geographical location~\cite{Lambiotte2008, Expert2011, Backstrom2010, Scellato2011, Illenberger2013}.

Segregation statistics, also referred to as segregation measures, are often used to quantify homophily~\cite{Rodriguez-Moral2016, Bojanowski2014}. Many approaches are based on co-presence in organisational units such as schools~\cite{Orfield2014}, voluntary associations~\cite{Popielarz1999}, occupations~\cite{Charles1995}, or census tracts~\cite{Reardon2004}, and we refer to them as \emph{organisational} statistics. Typically, they compare how the distribution of demographic attributes within organisational units differs from the distribution of attributes in the general population. Whilst organisational statistics are applicable whenever data can be stratified according to a variable of interest, they cannot capture segregation at smaller scales than the strata~\cite{Blumenstock2013}. For example, the ethnic composition of a set of schools may be representative of the general population, indicating that there is no (organisational) segregation. But the social networks \emph{within} schools often exhibit strong ethnic homophily~\cite{Currarini2009,Moody2001}. Organisational statistics cannot capture such social segregation.

@@

Building on the successes of conditionally-independent edge models~\cite{Snijders2011} and, in particular, latent space models for social networks~\cite{Hoff2002,Hoff2008}, we consider a generative model for social networks whose members occupy a multidimensional Blau space in \cref{sec:model}. We discuss desirable properties for social segregation statistics, and, using the generative network model, we develop a suite of statistics applicable to arbitrary attributes in \cref{sec:model-based-segregation}. The statistics capture segregation at different scales: single individuals, pairs of individuals, and society as a whole. Because of both their probabilistic foundations and their construction from the universal notion of a social tie, the segregation statistics have a universal scale, i.e.\ one unit of segregation has the same implications across different societies or at different times. We show that the segregation statistic for pairs of individuals can be a metric and can thus be used to quantify distance in Blau space. We illustrate the statistics with a simple example, and we show that it reduces to a well-known segregation statistic if the attributes are univariate and categorical: the natural logarithm of \citeauthor{Moody2001}'s $\alpha$ index~\cite{Moody2001}.

@@

We consider a generative model for social networks for a population of $n$ individuals $\population$ who occupy a Blau space $\blauspace$ spanned by their demographic attributes, such as age, income, or sex. In contrast to common latent space models~\cite{Hoff2002, Hoff2008}, the attributes are observed, although \textcite{Hoff2002} also consider an extension including covariates. The $q$-dimensional attribute vector $x_i\in\blauspace$ for each individual $i\in \population$ is drawn independently from a distribution $P(x_i)$ of demographic attributes. Elements of the attribute vector can take continuous, ordinal, or categorical values. Connections between individuals are encoded by the binary adjacency matrix $A$ such that $A_{ij}=1$ if $j$ considers $i$ to be a friend and $A_{ij}=0$ otherwise. We assume that people do not interact with themselves such that $A_{ii}=0$ for all $i$, and that connections are undirected, although social ties need not be reciprocated in general~\cite{Ball2013}.

@@

We have so far emphasised the desirability of metrics on social spaces with a universal characteristic scale (in the sense of being comparable between societies). After first developing universal social segregation statistics, including a notion of social separation, we will formulate both a metric and a notion of scale in \cref{sec:segregation-distance}.

@@

In other words, the social separation statistic is a true measure of \emph{distance} in the social space with a probabilistic interpretation if features are themselves measures of distance, including all the features we consider subsequently. This observation puts Peter Blau's~\cite{Blau1977} hypothesis that ``the macrostructure of societies can be defined as a multidimensional space of social positions among which people are distributed and which affect their social relations'' on a sound statistical footing: \emph{fitting conditionally-independent edge models allows us to learn the metric of Blau space}. The metric has a universal scale: \emph{one unit of social separation has the same probabilistic meaning independent of the society under consideration}, facilitating comparison across disparate datasets. Even if two societies have different Blau space dimensions, e.g.\ a society might exist that strongly discriminates based on characteristics which are not found in other societies, the social separation between a pair of individuals has a common meaning. % However, it is worth noting that we cannot quantify segregation due to unobserved dimensions of the social space.

@@

The General Social Survey (GSS) is a nationally-representative face-to-face survey of non-institutionalised adults living in the US. Demographic attributes of seeds are collected regularly and include age, sex, ethnicity, religion, and education~\cite{Marsden1987,Marsden1988}. In 2004, respondents were asked about the demographic background of people ``with whom they discuss important matters'', which tends to elicit close ties~\cite{Marin2004}. We omit all nominees who are not considered to be friends or who are family. Some of the demographic attributes of seeds and nominees are missing because respondents did not know or refused to provide the information, and we drop dyads associated with individuals with one or more missing attributes, as shown in \cref{tbl:survey-sample-size}. Such a complete-case analysis can introduce biases if the data are not missing completely at random, but handling the missing data in a principled fashion would require us to develop a model for demographic attributes~\cite{Pigott2001}.

The coding of age and sex is consistent amongst seeds and nominees. We aggregate the detailed coding of ethnic and religious attributes of seeds to match the coding of nominees, as shown in \cref{tbl:survey-gss-coding}. Kernel features include the absolute age and ordinal education level difference as well as binary indicators for differences along the sex, ethnicity, and religion dimensions. For each demographic attribute, we define a feature for the logistic kernel in \cref{eq:logistic-kernel}, as shown in \cref{tbl:survey-gss-coding}. To standardise the features $f(x_i,x_j)$, we subtract their mean and divide non-binary features by twice their standard deviation~\cite{Gelman2008a}; binary features are not rescaled. The statistics are calculated with respect to a random sample of pairs of seeds. Feature standardisation allows us to compare kernel parameters more easily~\cite{Gelman2008a} and simplifies the formulation of priors: we use independent, weakly-informative Cauchy priors for the kernel parameters such that
\begin{equation*}
    P(\theta_l) \propto {\left[1+{\left(\frac{\theta_l}{\alpha_l}\right)}^2\right]}^{-1}.
\end{equation*}
Following \textcite{Gelman2008}, we chose the scale parameters $\alpha_l=2.5$ for $l>1$ to represent our weak prior belief that changing a feature by one standard deviation is unlikely to change the log odds by more than five: the independent Cauchy distributions regularise the kernel parameters by placing significant prior probability near zero, but their heavy tails allow for significant departures from zero should the data be in support of large parameters. We set $\alpha_1=10$ because the parameter $\theta_1$ associated with the constant bias term could change significantly depending on the population size~\cite[chapter 16]{Gelman2013}.

@@

To get a better understanding of Blau space and the metric induced by the connectivity kernel, we consider a sample $S$ of 1,000 respondents from the GSS and USS. For each sample, we compute the social separation between pairs of respondents to obtain a distance matrix
\begin{equation*}
    \hat\varphi_{ij} = \transpose{\hat\theta}\left(f(x_j, x_j)-f(x_i, x_j)\right),
\end{equation*}
where $\hat\theta$ is the posterior median of the kernel parameters discussed in \cref{sec:survey-gss,sec:survey-bhps-usoc}. We use multidimensional scaling to embed the respondents in a two-dimensional space~\cite{Borg1996}, as shown in \cref{fig:segregation-embedding}. Panels~(c) and~(d) show the two-dimensional embedding that best approximates the distance matrix in the high-dimensional social space (we omit contribution due to physical space for the USS to make the embeddings comparable).

@@

Panel~(c) and~(d) of \cref{fig:segregation-embedding} also show the social isolation $\phi$ experienced by individuals as a greyscale heat map which we obtained in two steps: first, we evaluated an estimate of the social isolation
\begin{equation*}
    \hat\phi_i = \frac{1}{\card{S}-1}\sum_{j\in S:j\neq i} \hat\varphi_{ij}.
\end{equation*}
Second, we applied Gaussian kernel smoothing to the social isolation in the embedding space. Respondents occupying the centre of society experience little isolation whereas individuals in the periphery are more isolated. For example, members of the ethnic majority experience an average social isolation of 4.53 (4.47--4.59 95\% credible interval) in the USS and 4.46 (4.10--4.84 95\% credible interval) in the GSS. In contrast, the average social isolation amongst ethnic minorities is 4.99 (4.91--5.07 95\% credible interval) and 5.16 (4.73--5.58 95\% credible interval): significantly higher than for the ethnic majority.

@@

Similar to the social separation in \cref{eq:social-separation-logistic}, the social strain can be broken down into components
\begin{equation}
    \Phi_l = \theta_l \int dx\ dy\ \left(f_l(y, y) - f_l(x, y)\right)\label{eq:segregation-contributions}
\end{equation}
because it is a linear functional of $\varphi$: each component contributes to the social strain in society. For each of the datasets, we evaluate an estimate of the contributions to the social strain
\begin{equation*}
    \hat\Phi_l = \frac{2\theta_l}{\card{\seeds}(\card{\seeds}-1)} \sum_{i < j\in\seeds:i\neq j} \left[f_l(x_j, x_j)-f_l(x_i, x_j)\right],
\end{equation*}
where the sum is over all distinct pairs of seeds $\seeds$. The contribution $\Phi_l$ quantifies the average social separation due to feature $l$, and it captures the effect of both the connectivity kernel $\rho(x,y)$ and the attribute distribution $P(x)$ in Blau space: neither is sufficient on its own to quantify segregation. Furthermore, $\Phi$ and its contributions in \cref{eq:segregation-contributions} can facilitate comparison across different datasets, as illustrated in \cref{fig:segregation-embedding}: they have an intuitive interpretation (the probability of edges decreases with increasing segregation) and universal scale (one unit of segregation has the same effect on edge probability).

@@

Even though we did not expect the kernel parameters to be consistent across countries, time, or even different surveys, people connected with one another in a surprisingly similar fashion across the different datasets (the BHPS and USS are longitudinal studies such that consistent parameter estimates are less surprising). Our observations, together with a study by \textcite{Mossong2008} finding that ``mixing patterns [\ldots] were remarkably similar across different European countries'', suggest that connectivity kernels for friendships vary little across societies and time. To test this hypothesis, further surveys should be conducted in a unified fashion to minimise the effects of question wording and how the survey is administered~\cite{Eagle2015}. In particular, such surveys should explore options to explicitly incentivise nominees to provide data about themselves~\cite{Biernacki1981}: seeds may not recall certain attributes, or nominees may deliberately portray themselves inaccurately~\cite{Bruch2016}. Questions regarding ethnicity should allow respondents to provide multiple answers such that people with mixed ethnic backgrounds can express their identity. Rather than asking respondents about potentially sensitive information, such as income, proxy information that is more readily available---and potentially more informative of how individuals interact with society---could be collected~\cite{Po2012}. Whenever possible, aggregation of attributes such as age into bins should be avoided because it limits the ability to infer kernel parameters~\cite{Hutcheon2010}, as we saw in \cref{sec:survey-alp}. Connectivity kernels should be inferred jointly for all dimensions of Blau space to control for social preferences on correlated attributes.

@@

We would like to thank Sahil Loomba for useful feedback on the manuscript and acknowledge support from the grant EP/N014529/1.

\printbibliography
\end{refsection}

\ifdefined\composite
\pagebreak
\noindent{\titlefont\huge Supplementary material}

@@

\begin{refsection}
\section{Evaluation of the observed-data log-likelihood}

@@

In the first step, we draw the positions $x$ of $n=2,000$ nodes uniformly at random from the unit square, and we connect nodes to one another according to a logistic connectivity kernel with features
\begin{equation*}
    f(x_i, x_j) = \left(1, \frac{3 \left[\abs{x_{i1}-x_{j1}} - \frac{1}{3}\right]}{\sqrt{2}}, \frac{3 \left[\abs{x_{i2}-x_{j2}} - \frac{1}{3}\right]}{\sqrt{2}}\right),
\end{equation*}
where the first feature represents the bias, and the last two features capture distance in Blau space. The features were chosen to be standardised in the same fashion as described in \cref{sec:survey-gss}, i.e.\ to have zero mean and a standard deviation of $0.5$~\cite{Gelman2008a}. The corresponding parameter values $\theta$ are drawn from a normal distribution with unit variance and expectation
\begin{equation*}
    \left\langle \theta\right\rangle = \left(-7, 0, 0\right).
\end{equation*}
The expectation $\langle\theta\rangle$ was chosen such that it gives rise to a typical degree of $2,000 \times \expit(-7)\approx 1.8$, similar to the real-world datasets considered in \cref{sec:application}. We select $s=100$ respondents as egos and include all their alters as positive examples. We sample three times as many negative examples by selecting distinct pairs of respondents. If the number of respondents is not sufficient to draw the desired number of distinct control pairs, we use all $\frac{s(s-1)}{2}$ possible distinct pairs of respondents as negative examples.

@@

In the USS, respondents could identify with mixed ethnicities, and we coded such responses as a mixed membership. For example, a respondent who indicated ``mixed Asian and White'' would belong to both White and Asian ethnicities. To quantify how different two people are in terms of ethnicity, we define the feature map
\begin{equation*}
    f_\text{ethnicity}(x_i, x_j) = \frac{1}{2} \sum_{l\in E}\abs{x_{il}-x_{jl}},
\end{equation*}
where $E$ is the set of attributes encoding ethnic identity, and ethnicity memberships are normalised such that $\sum_{l\in E}x_{jl}=1$ for all $j$. For example, $f_\text{ethnicity}(x_i, x_j)=1$ for two people $i$ and $j$ one of which identifies as white and the other as black. For a person $i$ identifying as black and another person $j$ identifying as mixed black and Asian, $f_\text{ethnicity}(x_i, x_j)=0.5$.

@@

\printbibliography
\end{refsection}
\fi
